---
output: 
  revealjs::revealjs_presentation:
    self_contained: false
    reveal_plugins: ["notes"]
    theme: moon
    highlight: tango
    center: true
---

<section class="present">
<h1 class="title">Creating custom htmlwidgets for shiny</h1>
<h2 class="author">Mark Nielsen</h2>
<aside class="notes">
- Introduce yourself: USU, BYU, Merrick, now IHC.
- SAS, R, Web Development, finally shiny.
- htmlwidgets a mystery, started to understand, previously used `rCharts`.
- Training Garrett Grolemund recommended that I look into `htmlwidgets`.
</aside>
</section>

##

<section class="titleslide level1">
<h1>Setting Up the HTMLWidget</h1>

```{r, echo=TRUE, eval=FALSE}
devtools::create("mywidget")
setwd("mywidget")
htmlwidgets::scaffoldWidget("mywidget")
devtools::install()
```
To learn more, see Ramnath V., Kenton R., and Rstudio's [tutorial](http://www.htmlwidgets.org/develop_intro.html) on creating htmlwidgets.

<aside class="notes">
- Ready to create your first htmlwidget?
- Ramnath Vaidyanathan (rCharts), Kenton Russell, and RStudio
- Increadible tutorial.
</aside>

## File Structure
<pre><code>
.
├── DESCRIPTION
├── inst
│   └── htmlwidgets
│       ├── mywidget.js
│       └── mywidget.yaml
├── mywidget.Rproj
├── NAMESPACE
└── R
    └── mywidget.R
</code></pre>

<aside class="notes">
- <b>DEMO:</b> Look at file structure in RStudio
</aside>

## R output
```{r, eval=TRUE, echo=TRUE}
library(mywidget)
mywidget("Hello World")
```

<aside class="notes">
- Vioala! Your first htmlwidget AND the classic "Hello World".
- Pat yourself on the back.
- We can go home!
- <b>DEMO:</b> Let's dig into the code.
</aside>

</section>

##

<section class="titleslide level1">
<h1>Adding the JS Code</h1>

1. Find or develop the JS code you want to bind to R.
2. Replace JS code in `./inst/htmlwidgets/hive.js`
3. Copy supporting JS and CSS into `./inst/htmlwidgets/lib/` folder.
4. Define dependencies in `./inst/htmlwidgets/hive.yaml`

<aside class="notes">
- Okay, maybe not so cool, but we can do better.
- Find some JS code (I'm not a Web Developer)
</aside>

## Enter hive plots
- Like the functionality and look of Mike Bostock's D3 implementation of [hive plots](https://bost.ocks.org/mike/hive/).
- A simpler version is found [here](https://bl.ocks.org/mbostock/2066415).

<aside class="notes">
- Martin Krzysinski credited with hive plot.
- <b>DEMO:</b> Pull code up in R and show how I split it up
</aside>

</section>

##

<section class="titleslide level1">
<h1>Create the Bindings</h1>
Goal: Get R data to look just like this.
<pre>
<code class="hljs" data-trim>
var nodes = [
  {x: 0, y: .1},
  {x: 0, y: .9},
  {x: 1, y: .2},
  {x: 1, y: .3},
  {x: 2, y: .1},
  {x: 2, y: .8}
];
var links = [
  {source: nodes[0], target: nodes[2]},
  {source: nodes[1], target: nodes[3]},
  {source: nodes[2], target: nodes[4]},
  {source: nodes[2], target: nodes[5]},
  {source: nodes[3], target: nodes[5]},
  {source: nodes[4], target: nodes[0]},
  {source: nodes[5], target: nodes[1]}
];
</code>
</pre>

<aside class="notes">
- Now the tricky part... 
- Get my R dataframe to look like this d3 data object.
</aside>


## R Code
```{r, echo=TRUE, eval=FALSE}
hive <- function(nodes, 
                 links, 
                 innerRadius = 40, 
                 outerRadius = 240, 
                 opacity = 0.7, 
                 width = NULL, 
                 height = NULL, 
                 elementId = NULL) {

  # sort in order of node id
  if("id" %in% colnames(nodes)) {
    nodes <- nodes[order(nodes$id),]
    nodes$id <- NULL
  }

  # color by axis if no coloring is supplied
  if(!("color" %in% colnames(nodes))) {
    nodes$color <- nodes$x
  }

  # forward options using x
  x = list(
    nodes = nodes,
    links = links,
    numAxis = max(nodes$x)+1,
    options = list(innerRadius=innerRadius,
                   outerRadius=outerRadius,
                   opacity=opacity)
  )

  # create widget
  htmlwidgets::createWidget(
    name = 'hive',
    x,
    width = width,
    height = height,
    package = 'hiveD3',
    elementId = elementId
  )
}
```

<aside class="notes">
- `nodes` and `links` are R dataframes
- Function is passing list `x` to JS
- Note tweaks made to dataframes based on input.
</aside>

## htmlwidgets
<pre>
<code class="hljs" data-trim>
// alias options
var options = x.options;

// convert links and nodes data frames to d3 friendly format
var nodes = HTMLWidgets.dataframeToD3(x.nodes);
var prelinks = HTMLWidgets.dataframeToD3(x.links);

// create json of link sources and targets
var links = [];
prelinks.forEach(function(d){
  var tmp = {};
  tmp.source=nodes[d.source];
  tmp.target=nodes[d.target];
  links.push(tmp);
});
</code>
</pre>

<aside class="notes">
- Okay, not so bad.
- Recreating nodes is easy.
- Don't want to touch js code much... make dat exact
- Need to loop through each prelink and create links
</aside>

</section>

##

<section class="titleslide level1">
<h1>Putting it all together</h1>

```{r, echo=TRUE, eval=FALSE}
library(hiveD3)
nodes = data.frame(id=c(0,1,2,3,4,5,6,7,8),
                   x=c(0,0,1,1,2,2,3,3,4), 
                   y=c(.1,.9,.2,.3,.1,.8,.3,.5,.9))
links = data.frame(source=c(0,1,2,2,3,4,5,6,7,8,8),
                   target=c(2,3,4,5,5,6,7,8,8,0,1))


hive_no_int(nodes=nodes,links=links, width = "100%", height = "500px")
```

<aside class="notes">
- All my bindings are set-up.
- I've built and loaded my package.
- Define some dataframes and we're ready to test it out.
</aside>

## Non-Interactive R Output
```{r, eval=TRUE, echo=FALSE}
library(hiveD3)
nodes = data.frame(id=c(0,1,2,3,4,5,6,7,8),
                   x=c(0,0,1,1,2,2,3,3,4), 
                   y=c(.1,.9,.2,.3,.1,.8,.3,.5,.9))
links = data.frame(source=c(0,1,2,2,3,4,5,6,7,8,8),
                   target=c(2,3,4,5,5,6,7,8,8,0,1))


p1 <- hive_no_int(nodes=nodes,links=links, width = "100%", height = "500px")
p1
```

<aside class="notes">
- Cool, but first question is, why can't I interact with it?
- So I do have to do some JS coding anyway.
</aside>
</section>

## Finishing Touches
- Adding interaction
- Creating and sharing your package
- Creating R documentation using RStudio and roxygen2
- Adding your package to [htmlwidget gallery](http://gallery.htmlwidgets.org/)

<aside class="notes">
- <b>DEMO:</b> Open R code, R-help to show documentation.
- <b>DEMO:</b> If time show JS code to add interactivity.
- Show that a lot htmlwidgets reference d3.
- Search for 'hive'... okay, not so original.
</aside>
##

<section class="titleslide level1">
<h1>The Final Product</h1>

```{r, echo=TRUE, eval=FALSE}
library(devtools)
install_github('nielsenmarkus11/hiveD3')

library(hiveD3)

nodes = data.frame(id=c(0,1,2,3,4,5,6,7,8),
                   x=c(0,0,1,1,2,2,3,3,4), 
                   y=c(.1,.9,.2,.3,.1,.8,.3,.5,.9))
links = data.frame(source=c(0,1,2,2,3,4,5,6,7,8,8),
                   target=c(2,3,4,5,5,6,7,8,8,0,1))

hive(nodes=nodes,links=links, width = "100%", height = "500px")
```
<aside class="notes">
- I've added interaction
- I've created documentation
- I've shared on GitHub
</aside>

## Interactive R Output
```{r, eval=TRUE, echo=FALSE}
library(rmarkdown)
library(hiveD3)

nodes = data.frame(id=c(0,1,2,3,4,5,6,7,8),
                   x=c(0,0,1,1,2,2,3,3,4), 
                   y=c(.1,.9,.2,.3,.1,.8,.3,.5,.9))
links = data.frame(source=c(0,1,2,2,3,4,5,6,7,8,8),
                   target=c(2,3,4,5,5,6,7,8,8,0,1))

p1 <- hive(nodes=nodes,links=links, width = "100%", height = "500px")
p1
```
<aside class="notes">
- Next steps:
- CRAN? (New package or integrate into networkD3)
- More interactivity and flexibility!
</aside>
</section>

##

<section class="titleslide level1">
<h1>References</h1>
- Bostock M, Morin R (2012). [Hive Plots](https://bost.ocks.org/mike/hive/). Retrieved from https://bost.ocks.org/mike/hive/.

- Bostock M (2016). [Hive Plot (Links)](https://bl.ocks.org/mbostock/2066415). Retrieved from https://bl.ocks.org/mbostock/2066415.

- Bostock M (2017). [D3 Data-Driven Documents](https://d3js.org/). Retrieved from https://d3js.org/.

## References (cont.)
- Krzywinski M, Birol I, Jones S, Marra M (2011). [Hive Plots — Rational Approach to Visualizing Networks](https://academic.oup.com/bib/article/13/5/627/412507/Hive-plots-rational-approach-to-visualizing). Briefings in Bioinformatics (early access 9 December 2011, doi: 10.1093/bib/bbr069).

- Vaidyanathan R, Russell K, RStudio, Inc. (2014-2015). [Creating a widget](http://www.htmlwidgets.org/develop_intro.html). Retrieved from http://www.htmlwidgets.org/develop_intro.html.
</section>

## Questions?
